---
title: "Leaflet: An introduction to creating maps with leaflet"
author: "Barry Bullas - Anthony Foster"
date: "25 May 2019"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

# Introduction
## What is Leaflet
Leaflet is the leading open-source JavaScript library for mobile-friendly interactive maps. Weighing just about 38 KB of JS, it has all the mapping features most developers ever need.

Leaflet is designed with simplicity, performance and usability in mind. It works efficiently across all major desktop and mobile platforms, can be extended with lots of plugins, has a beautiful, easy to use and well-documented API and a simple, readable source code that is a joy to contribute to.

Leaflet is one of the most popular open-source JavaScript libraries for interactive maps. It’s used by websites ranging from The New York Times and The Washington Post to GitHub and Flickr, as well as GIS specialists like OpenStreetMap, Mapbox, and CartoDB.

This R package makes it easy to integrate and control Leaflet maps in R.

A cheatsheet for this package is here: https://ugoproto.github.io/ugo_r_doc/leaflet-cheat-sheet.pdf

```{r echo=FALSE, warning=FALSE}
#Example of a simple map

#Check if leaflet is installed, if not install it
require("leaflet")
require("DT")
require("readr")
library(leaflet)
library(DT)
library(readr)

# Add leaflet extras
require("leaflet.extras")
library(leaflet.extras)

# A simple map...

m <- leaflet() %>%
  addTiles() %>%  # Add default OpenStreetMap map tiles
  addMarkers(lng=174.768, lat=-36.852, popup="The birthplace of R")
m  # Print the map

```

## Features
* Interactive panning/zooming
* Compose maps using arbitrary combinations of:
* Map tiles
* Markers
* Polygons
* Lines
* Popups
* GeoJSON

## Guidance
You can get more information on the using the Leaflet library in r from the [Leaflet Github Repository](https://rstudio.github.io/leaflet/).  
You can get more information on Leaflet from the [Leaflet website](https://leafletjs.com/index.html).
[This github](https://andrewbtran.github.io/NICAR/2017/maps/leaflet-r.html#how_to_put_the_map_online) also has some helpful examples

***

# Building Maps
## Getting Started
The first thing to do when buiding leaflet maps in r is to install the leaflet package using install.packages, then attach it using the Library function.

```{r eval=FALSE}
install.packages("leaflet")
library(leaflet)
```

***

## Simple map
You create a Leaflet map with these basic steps:

Create a map widget by calling leaflet().

1. Add layers (i.e., features) to the map by using layer functions (e.g. addTiles, addMarkers,  addPolygons) to modify the map widget.
2. Repeat step 2 as desired.
3. Print the map widget to display it.

Here’s a basic example:
```{r}
m <- leaflet() %>%
  # Add default OpenStreetMap map tiles
  addTiles() %>%  
  # add a marker for Buckhingham palace
  addMarkers(lng=-0.1419, lat=51.5014, popup="Buckingham Palace, London")
m  # Print the map
```

***

### Base Map Tiles
The easiest way to add tiles is by calling addTiles() with no arguments; by default, OpenStreetMap tiles are used.

There are a range of popular free third-party basemaps which can be added using the addProviderTiles() function, which is implemented using the leaflet-providers plugin. You can [view examples of each base map  here](http://leaflet-extras.github.io/leaflet-providers/preview/index.html).

As a convenience, leaflet also provides a named list of all the third-party tile providers that are supported by the plugin. Just type providers$ and choose from one of the options. You can also use names(providers) to view all of the options.

The example below uses the ESRI World Top Map as a basemap.  It also uses the setView function to set the intial position and zoom level of the map.  If you do not call the setView function, the default view will include any markers you have placed on the map.  As the example below does not have any markers, I have used setView to focus on a specific area rather than the intial view showing the whole world.

setView allows you to specify a point on the map that you want your leaflet map to centre on.  You can get centre point coordinates, including box boundary coordinates, in many different CRS here: http://bboxfinder.com

```{r}
m <- leaflet() %>%
  #Use the Esri World Topo Map as the basemap
  addProviderTiles(providers$Esri.WorldTopoMap)%>%
  # Setview sets the intial position of the map to Scotland
  setView(lng = -5.394, lat = 56.821, zoom = 6)

# Print the map
m

```

***

## Loading Markers from a file
This example loads the location of the 32 NFL franchises and plots them on a map.

The data is in csv format and contains four columns:

1. Team - The name of the NFL franchise
2. Confrence - AFC or NFC
3. Latitude - The latitude of the franchise
4. Longitude - The longitude of the franchise

You can explore the data here

```{r echo=FALSE, warning=FALSE}
#load the NFL data from a csv
teams <- read_csv("data/nfl_franchises.csv")

# Load the DataTables package
library(DT)

# Show the data using the datatable function
datatable(teams)

```

By loading the data from the csv and passing it to leaflet we can display it as markers on the map.

```{r}
#Load the NFL data from the csv file
teams <- read_csv("data/nfl_franchises.csv")

m <- leaflet(teams) %>%
  # Add default OpenStreetMap map tiles
  addTiles() %>%
  # add a marker for each team using the TeamName variable as
  addMarkers(~Lon, ~Lat, popup = ~TeamName)
m  # Print the map
```

***

## Heat map
RLeaflet is a basic leaflet package that can be augmented with additional functions from the "leaflet.extras" packages.  Installing this package will add a lot of extra features to the leaflet package, such as making a heatmat of your data.

The leaflet.extras package is already installed, so lets view the 'quakes' dataset that is included in RStudio;

```{r echo=FALSE}
head(quakes)
```

As our data has WGS84 lat/long data, leaflet can project it without any issues.  Using the leaflet.extras fucntion "addWebGLHeatmap" we can display the quakes data in a heatmap.  To make the best use of the heatmap we should use a basemap that complements it.  See https://leaflet-extras.github.io/leaflet-providers/preview/ for a list of available basemaps for leaflet.extras.

We'll use the ESRI WorldImagery map and set the intensity of the heat according to the magnitude of the measured earthquakes to get an idea of where the quakes are at their worst;

```{r echo=TRUE}

leaflet(quakes) %>% addProviderTiles(providers$Esri.WorldImagery) %>%
  setView( 178, -20, 5 ) %>%
    addHeatmap(lng = ~long, lat = ~lat, intensity = ~mag,
               blur = 25, max = 0.05, radius = 12)

```


***

## Cluster Map

RLeaflet can 'cluster' points together to provide an aggregated view.  This can be quite hard to understand without the aid of an example, so let's use UK accidents data from 2017.  

This is freely available data that provides details of approximately 130,000 accident locations across the UK, including speed, number of vehicles involved and several more variables.  A csv of the data is available in this repository.

```{r warning=FALSE}
uk_acc_2017 <- read_csv("uk_accidents_2017.csv")
#remove missing values from lat/long
uk_acc_2017 <- uk_acc_2017[complete.cases(uk_acc_2017[, 4:5]),]

head(uk_acc_2017[,4:12])

```

Lets do a simple plot of the points on a map to see how the raw data looks;

```{r echo=FALSE}
leaflet(uk_acc_2017) %>%
  # Use OpenStreetMap as the basemap
  addProviderTiles(providers$OpenStreetMap.Mapnik)%>%
  setView(lng = -4, lat = 54, zoom = 7) %>%
  # Add clustered points to the map
  addCircles(radius = 3, opacity = 0.75)


```

Striking but not very easy to interpret.  There are so many points that much of the land is completely covered, making it very difficult to understand how accidents are spread across the country.

Next, lets add the clustering function to aggregate some of those points together;

```{r echo=TRUE}
leaflet(uk_acc_2017) %>%
  # Use OpenStreetMap as the basemap
  addProviderTiles(providers$OpenStreetMap.Mapnik)%>%
  setView(lng = -4, lat = 54, zoom = 7) %>%
  # Add clustered points to the map
  addMarkers(clusterOptions = markerClusterOptions() )
```

Try zooming in and out.  Notice how the clustering is amended depending on zoom level?  Clustering makes the data easier to interpret.  Hover over one of the numbered points and you will see the outline of the area that it represents.


***

## Choropleth Map
```{r}

```

***

# Other features
## Layer Groups and Layers Control
```{r}

```

***

## Base Maps
```{r}

```

***

## Custom Markers
```{r}

```

***

## Popups
```{r}

```

***

## Zoom Levels
```{r}

```

***

## Mini Maps
```{r}

```
