---
title: "Leaflet: An introduction to creating maps with leaflet"
author: "Barry Bullas"
date: "25 May 2019"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

# Introduction
## What is Leaflet
Leaflet is the leading open-source JavaScript library for mobile-friendly interactive maps. Weighing just about 38 KB of JS, it has all the mapping features most developers ever need.

Leaflet is designed with simplicity, performance and usability in mind. It works efficiently across all major desktop and mobile platforms, can be extended with lots of plugins, has a beautiful, easy to use and well-documented API and a simple, readable source code that is a joy to contribute to.

Leaflet is one of the most popular open-source JavaScript libraries for interactive maps. It’s used by websites ranging from The New York Times and The Washington Post to GitHub and Flickr, as well as GIS specialists like OpenStreetMap, Mapbox, and CartoDB.

This R package makes it easy to integrate and control Leaflet maps in R.
```{r echo=FALSE, warning=FALSE}
#Example of a simple map

#Check if leaflet is installed, if not install it
list.of.packages <- c("leaflet", "DT")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)

#Load Leaflet package
library(leaflet)

m <- leaflet() %>%
  addTiles() %>%  # Add default OpenStreetMap map tiles
  addMarkers(lng=174.768, lat=-36.852, popup="The birthplace of R")
m  # Print the map

```

## Features
* Interactive panning/zooming
* Compose maps using arbitrary combinations of:
* Map tiles
* Markers
* Polygons
* Lines
* Popups
* GeoJSON

## Guidance
You can get more information on the using the Leaflet library in r from the [Leaflet Github Repository](https://rstudio.github.io/leaflet/).  
You can get more information on Leaflet from the [Leaflet website](https://leafletjs.com/index.html){:target="_blank"}.
[This github](https://andrewbtran.github.io/NICAR/2017/maps/leaflet-r.html#how_to_put_the_map_online){:target="_blank"} also has some helpful examples

***

# Building Maps
## Getting Started
The first thing to do when buiding leaflet maps in r is to install the leaflet package using install.packages, then attach it using the Library function.

```{r eval=FALSE}
install.packages("leaflet")
library(leaflet)
```

***

## Simple map
You create a Leaflet map with these basic steps:

Create a map widget by calling leaflet().

1. Add layers (i.e., features) to the map by using layer functions (e.g. addTiles, addMarkers,  addPolygons) to modify the map widget.
2. Repeat step 2 as desired.
3. Print the map widget to display it.

Here’s a basic example:
```{r}
m <- leaflet() %>%
  # Add default OpenStreetMap map tiles
  addTiles() %>%  
  # add a marker for Buckhingham palace
  addMarkers(lng=-0.1419, lat=51.5014, popup="Buckingham Palace, London") 
m  # Print the map
```

***

### Base Map Tiles
The easiest way to add tiles is by calling addTiles() with no arguments; by default, OpenStreetMap tiles are used.

There are a range of popular free third-party basemaps which can be added using the addProviderTiles() function, which is implemented using the leaflet-providers plugin. You can [view examples of each base map  here](http://leaflet-extras.github.io/leaflet-providers/preview/index.html){:target="_blank"}.

As a convenience, leaflet also provides a named list of all the third-party tile providers that are supported by the plugin. Just type providers$ and choose from one of the options. You can also use names(providers) to view all of the options.

The example below uses the ESRI World Top Map as a basemap.  It also uses the setView function to set the intial position and zoom level of the map.  If you do not call the setView function, the default view will include any markers you have placed on the map.  As the example below does not have any markers, I have used setView to focus on a specific area rather than the intial view showing the whole world.

```{r}
m <- leaflet() %>%
  #Use the Esri World Topo Map as the basemap
  addProviderTiles(providers$Esri.WorldTopoMap)%>%
  # Setview sets the intial position of the map to Scotland
  setView(lng = -5.394, lat = 56.821, zoom = 6) 

# Print the map
m

```

***

## Loading Markers from a file
This example loads the location of the 32 NFL franchises and plots them on a map.

The data is in csv format and contains four columns:

1. Team - The name of the NFL franchise
2. Confrence - AFC or NFC
3. Latitude - The latitude of the franchise
4. Longitude - The longitude of the franchise

You can explore the data here

```{r echo=FALSE, warning=FALSE}
#load the NFL data from a csv
teams <- read.csv("data/nfl_franchises.csv", stringsAsFactors=F)

# Load the DataTables package
library(DT)

# Show the data using the datatable function
datatable(teams)

```

By loading the data from the csv and passing it to leaflet we can display it as markers on the map.

```{r}
#Load the NFL data from the csv file
teams <- read.csv("data/nfl_franchises.csv", stringsAsFactors=F)

#pass the dataset into leaflet
m <- leaflet(teams) %>%
  # Add the Esri World Gray Canvas base map to help make the team locations stand out
  addProviderTiles(providers$Esri.WorldGrayCanvas) %>%
  # Add circles to the map for the teams using the Lat and Lon fields.  
  #The popup for each circle is the team name
  addCircles(~Lon, ~Lat, popup=teams$Team, weight = 3, radius=40, color="#C83803", stroke = TRUE, fillOpacity = 1) 
m  # Print the map
```

***

## Heat map

```{r}

```

***

## Cluster Map
```{r}

```

***

## Choropleth Map
```{r}

```

***

# Other features
## Layer Groups and Layers Control
```{r}

```

***

## Base Maps
```{r}

```

***

## Custom Markers
```{r}

```

***

## Popups
```{r}

```

***

## Zoom Levels
```{r}

```

***

## Mini Maps
```{r}

```

